Reintroduction of resistant frogs facilitates landscape-scale recovery in the presence of a lethal fungal disease

### 2024-09-27 doc/manuscript/submission_files/translocation.tex
Use default cite method instead of natbib

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{blkarray}
\usepackage{lineno}\linenumbers
\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{\textbf{Fig.}}
\else
  \newcommand\figurename{\textbf{Fig.}}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\usepackage{bookmark}

@@

Human activities are devastating global biodiversity
\citep{ceballos2015}, with important implications for ecosystem
resilience and human welfare \citep{naeem2009}. One consequence of human
alteration of the biosphere is an increase in emerging infectious
diseases \citep{jones2008, fisher2012}. Such diseases pose a severe
threat to wildlife populations \citep{daszak2000}, and have caused
dramatic declines and extinctions in a wide range of taxa, including
echinoderms, mammals, birds, and amphibians
\citep{hewson2014, samuel2015, scheele2019, cunningham2021}. Amphibians
are experiencing particularly devastating impacts of disease due to the
recent emergence and global spread of the highly virulent amphibian
chytrid fungus, \emph{Batrachochytrium dendrobatidis} (Bd)
\citep{luedtke2023, scheele2019}. By one estimate, hundreds of species
have experienced Bd-caused declines, and numerous susceptible taxa are
extinct in the wild \citep{scheele2019}. The escalating frequency of
disease emergence and the strength of impacts on wild populations
increasingly highlight the critical need for successful mitigation of
disease in species conservation.

Following pathogen arrival in a host population, resistance (ability to
limit pathogen burden) and tolerance (ability to limit the harm caused
by a particular burden) are key mechanisms by which hosts reduce disease
impacts \citep{raberg2009}. The effectiveness of these mechanisms has
direct consequences for the persistence and recovery of host populations
in the presence of disease \citep{brannelly2021}. Host immunity and
evolution both play important roles in the development of resistance and
tolerance, and are key targets for strategies aimed at mitigating
disease impacts in the wild \citep{garner2016}.

Evidence of natural recovery in the many Bd-impacted amphibian
populations is surprisingly rare (for notable exceptions, see
\citep{scheele2017, voyles2018, knapp2016}). This apparent low
resilience to disease may be due to the limited ability of many
amphibians to develop Bd resistance or tolerance via either immunity
\citep{rosenblum2012, fites2013, grogan2018a} or evolution (but see
\citep{savage2016, grogan2018b}). In turn, this would also reduce the
effectiveness of potential Bd mitigation strategies. For example, a
limited ability by amphibians to develop resistance or tolerance
suggests that reintroduction of amphibians into sites to reestablish
populations extirpated by Bd will often result not in population
recovery, but instead in the rapid reinfection and mortality of the
introduced animals and/or their progeny
\citep{hammond2021, knapp2022, stockwell2008, scheele2021}. If true, the
future of many amphibian species threatened by Bd appears bleak.

The Sierra Nevada yellow-legged frog (\emph{Rana sierrae}), is
emblematic of the global decline of amphibians caused by Bd
\citep{scheele2019}. Once the most common amphibian in the high
elevation portion of California's Sierra Nevada mountains, USA
\citep{grinnell1924}, during the past century this frog has disappeared
from more than 90\% of its historical range \citep{vredenburg2007}. Due
to the severity of its decline and the increasing probability of
extinction, \emph{R. sierrae} is now listed as ``endangered'' under the
U.S. Endangered Species Act. The decline of \emph{R. sierrae} was
initiated by the introduction of non-native trout into the extensive
fishless region of the Sierra Nevada \citep{bradford1989, knapp2000}
starting in the late 1800s. The arrival of Bd in the mid-1900s and its
subsequent spread \citep{vredenburg2019} caused additional large-scale
population extirpations \citep{vredenburg2010, rachowicz2006}. These
Bd-caused declines are fundamentally different from the fish-caused
declines because fish eradication is feasible \citep{knapp1998} and
results in the rapid recovery of frog populations
\citep{knapp2007, vredenburg2004}. In contrast, Bd appears to persist in
habitats even in the absence of amphibian hosts \citep{walker2007}, and
therefore represents a long-term alteration of invaded ecosystems that
amphibians will need to overcome to reestablish populations.

Despite the catastrophic impact of Bd on \emph{R. sierrae}, wherein most
Bd-naive populations are extirpated following Bd arrival
\citep{vredenburg2010}, some populations have persisted after epizootics
\citep{briggs2010} and are now recovering \citep{knapp2016}
(Fig.~\ref{fig-recovery-model}). During the epizootic phase, \emph{R.
sierrae} are highly susceptible to Bd infection as indicated by very
high infection intensities (``load''). In contrast, frogs in recovering
populations have reduced susceptibility to Bd infection
\citep{knapp2016}, with loads on adults consistently in the
low-to-moderate range \citep{briggs2010, knapp2011, joseph2018}. This
reduced susceptibility is evident even under controlled laboratory
conditions \citep{knapp2016}, indicative of host resistance against Bd
infection and not simply an effect of factors external to individual
frogs (e.g., environmental conditions). In addition to frogs from
recovering populations having higher resistance to Bd infection than
those from naive populations, they could also have higher tolerance, but
no data are currently available to evaluate this possibility. Therefore,
we focus on resistance throughout this paper.

The observed resistance of \emph{R. sierrae} could be the result of
several mechanisms that are not mutually exclusive, including natural
selection for more resistant genotypes \citep{savage2016, grogan2018b},
acquired immunity \citep{grogan2018a}, and/or inherent
between-population differences that pre-date Bd exposure. The possible
evolution of resistance and subsequent population recovery is consistent
with that expected under ``evolutionary rescue'', whereby rapid
evolutionary change increases the frequency of adaptive alleles and
restores positive population growth \citep{carlson2014, searle2020}.
This intriguing possibility also suggests an opportunity to expand
recovery beyond the spatial scale possible under natural recovery by
using resistant frogs from recovering populations in reintroductions to
vacant habitats \citep{joseph2018, mendelson2019}
(Fig.~\ref{fig-recovery-model}).

@@

Bd loads were fairly consistent before versus after translocation, and
loads were nearly always well below the level indicative of severe
chytridiomycosis (i.e., the disease caused by Bd) and associated frog
mortality \citep{joseph2018, vredenburg2010} (Supplementary Fig. 1).
Although it is possible that the observed relatively small changes in
load are a consequence of individuals with high Bd loads dying and
therefore being unavailable for sampling during the post-translocation
period, the fact that there was little difference in pre- versus
post-translocation Bd loads even in those populations that had very high
frog survival (70556, 74976 - see below; Supplementary Fig. 1) suggests
a true lack of substantial change in Bd load.

The ultimate measure of reintroduction success is the establishment of a
self-sustaining population. Given that it can take years or even decades
to determine the self-sustainability of a reintroduced population (for
an example in \emph{R. sierrae}, see \citep{joseph2018}), the use of
proxies is essential for providing shorter-term insights into
reintroduction success and the factors driving it. Results from our CMR
surveys allowed us to accurately estimate frog survival, including over
the entire CMR time series for each site and during only the 1-year
period immediately following translocation. These estimates were made
using site-specific models analyzed using the mrmr package
\citep{joseph2019} (\url{https://snarl1.github.io/mrmr/index.html}). We
use these estimates to describe general patterns of frog survival in all
translocated cohorts, and in an among-site meta-analysis of frog
survival to identify important predictors of 1-year frog survival (e.g.,
Bd load).

@@

Results from the frog translocation study are suggestive of population
establishment. However, a decade or more of surveys may be necessary to
confirm that populations are in fact self-sustaining \citep{joseph2018}.
To extend our inferences of population establishment beyond those
possible from the site-specific CMR data, we developed a population
viability model. Specifically, to test whether the observed yearly adult
survival probabilities in translocated populations were sufficient for
long-term viability, we built a stage-structured matrix model that
captured known frog demography and included demographic and
environmental stochasticity. We parameterized the model using CMR data
from translocated populations and known life history values in this
system (Supplementary Table 1). Results from a sensitivity analysis are
provided in Supplementary Fig. 4.

@@

Disease-induced population declines are decimating global biodiversity
\citep{daszak2000}, but broadly-applicable strategies to recover
affected species are generally lacking (e.g., \citep{garner2016}). Here,
we tested the possibility that populations of resistant individuals from
naturally recovering populations can be used to reestablish extirpated
populations of the endangered \emph{R. sierrae} in the presence of a
highly virulent pathogen (Bd). Our results indicate (i) the capacity of
reintroduced populations to become established and eventually recover
despite ongoing disease, (ii) that post-translocation frog survival is
influenced by site, cohort, and individual-level characteristics, and
(iii) that 50\% of the reintroduced populations are likely to persist
over a 50-year period. In light of the generally low success rate of
amphibian reintroduction efforts \citep{dodd2005}, our success in
reestablishing \emph{R. sierrae} populations via reintroduction of
resistant individuals is striking, and even more so given that \emph{R.
sierrae} was driven to near-extinction by Bd. Collectively, these
results provide a rare example of amphibian recovery in the presence of
Bd, and have important implications for the conservation and recovery of
amphibians and other taxa worldwide that are endangered by escalating
impacts from emerging infectious diseases.

Previous field studies in \emph{R. sierrae} show that frog-Bd dynamics
and frog survival in the presence of Bd are fundamentally different
between naive and recovering populations. Following the arrival and
establishment of Bd in previously-naive populations, adult frogs develop
high Bd loads that lead to mass die-offs \citep{vredenburg2010}. In
contrast, in recovering populations adult frogs typically have
low-to-moderate and relatively constant Bd loads and mass die-offs are
not observed \citep{briggs2010, knapp2011} (see also Supplementary Fig.
2). The differences in Bd load of frogs from naive and recovering
populations are also observed in controlled laboratory studies (see
Figure 4 in \citep{knapp2016}), and clearly indicate that frogs from
recovering populations exhibit resistance against Bd infection. Finally,
reintroductions of \emph{R. sierrae} collected from Bd-naive populations
and reintroduced into Bd-positive sites have consistently failed due to
the development of high Bd loads and resulting high frog mortality,
suggesting the importance of frog resistance in population
reestablishment in the presence of Bd.

@@

Results from the population viability model were also encouraging. In
particular, translocated populations with \textgreater{} 50\% survival
in the first year post-translocation were predicted to have a low
probability of extinction over 50 years (probability of extinction \(<\)
0.5 when year-1 juvenile survival probability was greater than 0.10).
The viability model highlighted the important role of frog survival in
affecting long-term population viability, and allowed us to extend the
temporal scale of our study beyond the years covered by our
post-translocation surveys. These long-term forecasts are important,
given that reintroduced \emph{R. sierrae} populations may often take
decades to achieve our ultimate goal of self-sustainability
\citep{joseph2018}. Making well-supported projections about the
long-term outcome of reintroduction efforts from shorter-term
information is crucial for adaptive management of species reintroduction
programs \citep{seddon2007}, including the one we are carrying out for
\emph{R. sierrae}. Specifically, the combined results from our
reintroduction study and viability modeling suggest that survival of
frogs in the first year following translocation is an effective proxy of
longer-term survival and population viability. In addition, given the
repeatability of frog survival at a site, 1-year frog survival also
serves as an effective proxy of site quality (i.e., the ability of a
site to support high frog survival and a viable frog population over the
long term). This proxy of site quality is important in the \emph{R.
sierrae} system because accurately predicting the ability of a site to
support a viable frog population \emph{a priori} remains difficult, even
after conducting 24 translocations over 16 years.

@@

Despite the demonstrated resistance of adult \emph{R. sierrae} against
Bd infection, individual and population-level impacts of Bd are still
evident. In an earlier study of 2 of our 12 translocated populations
\citep{joseph2018}, Bd infection and load had detectable effects on the
survival of adults and may have influenced population establishment
(sites referred to as ``Alpine'' and ``Subalpine'' in \citet{joseph2018}
are identified as ``70550'' and ``70505'' in the current study).
Applying similar analyses to all 12 of our translocated populations
would likely provide a broader perspective of the ongoing effect of Bd.
In addition to these important but relatively subtle effects of Bd on
adults, the impacts on younger life stages are more apparent. \emph{R.
sierrae} immediately following metamorphosis (``metamorphs'') are highly
susceptible to Bd infection \citep{ellison2019} and as a result
experience high mortality \citep{rachowicz2006}. This high
susceptibility of metamorphs is documented in numerous species of
anurans, and may result from the poorly developed immune system
characteristic of this life stage \citep{humphries2022}. In both
naturally recovering and reintroduced \emph{R. sierrae} populations, we
suggest that the high mortality of metamorphs is an important limitation
on subsequent recruitment of new adults. Therefore, although adult
\emph{R. sierrae} appear relatively resistant, Bd infection continues to
have important limiting effects on recovering populations (see also
\citep{hollanders2022}). The ongoing high susceptibility of metamorphs
to chytridiomycosis also highlights the critical need to develop better
methods of monitoring this vulnerable life stage, the results of which
could significantly advance the conservation and recovery of Bd-impacted
amphibians.

Our successful use of translocations to reestablish \emph{R. sierrae}
populations despite ongoing Bd infection is particularly important
because natural recovery alone appears insufficient to allow
reestablishment of populations at the landscape scale. Naturally
recovering \emph{R. sierrae} populations in Yosemite are relatively rare
and are often isolated from other such populations and from suitable but
vacant habitats. This isolation is the consequence of both natural
fragmentation due to rugged geography and the ongoing presence of
introduced, predatory trout in the majority of lakes and streams
\citep{bradford1993}. Collectively, these factors often prevent
naturally recovering \emph{R. sierrae} populations from expanding via
dispersal of frogs to nearby unoccupied habitats. Translocations allowed
us to overcome these dispersal limitations and reestablish populations
in sites that would otherwise remain unoccupied for the foreseeable
future. Of particular note are the translocations that we conducted as
part of the current study into areas from which \emph{R. sierrae} were
completely extirpated, but instead of single isolated habitats these
areas contained habitats composed of interconnected fishless lakes,
ponds, marshes, and/or streams. Successful reestablishment of one
\emph{R. sierrae} population in such areas has produced a source of
frogs from which adults and juveniles are now dispersing into adjacent
habitats. This process of translocation, establishment, and subsequent
dispersal has significant potential to eventually allow the
reestablishment of robust metapopulations that may be more resilient to
current and future natural and anthropogenic impacts than more
fragmented populations (e.g., \citep{heard2015}).

>>>>

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% journal abbreviations
\usepackage{jabbrv}
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{blkarray}
\usepackage{lineno}\linenumbers
\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{\textbf{Fig.}}
\else
  \newcommand\figurename{\textbf{Fig.}}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

Human activities are devastating global biodiversity\textsuperscript{1},
with important implications for ecosystem resilience and human
welfare\textsuperscript{2}. One consequence of human alteration of the
biosphere is an increase in emerging infectious
diseases\textsuperscript{3,4}. Such diseases pose a severe threat to
wildlife populations\textsuperscript{5}, and have caused dramatic
declines and extinctions in a wide range of taxa, including echinoderms,
mammals, birds, and amphibians\textsuperscript{6--9}. Amphibians are
experiencing particularly devastating impacts of disease due to the
recent emergence and global spread of the highly virulent amphibian
chytrid fungus, \emph{Batrachochytrium dendrobatidis}
(Bd)\textsuperscript{8,10}. By one estimate, hundreds of species have
experienced Bd-caused declines, and numerous susceptible taxa are
extinct in the wild\textsuperscript{8}. The escalating frequency of
disease emergence and the strength of impacts on wild populations
increasingly highlight the critical need for successful mitigation of
disease in species conservation.

Following pathogen arrival in a host population, resistance (ability to
limit pathogen burden) and tolerance (ability to limit the harm caused
by a particular burden) are key mechanisms by which hosts reduce disease
impacts\textsuperscript{11}. The effectiveness of these mechanisms has
direct consequences for the persistence and recovery of host populations
in the presence of disease\textsuperscript{12}. Host immunity and
evolution both play important roles in the development of resistance and
tolerance, and are key targets for strategies aimed at mitigating
disease impacts in the wild\textsuperscript{13}.

Evidence of natural recovery in the many Bd-impacted amphibian
populations is surprisingly rare (for notable exceptions,
see\textsuperscript{14--16}). This apparent low resilience to disease
may be due to the limited ability of many amphibians to develop Bd
resistance or tolerance via either immunity\textsuperscript{17--19} or
evolution (but see\textsuperscript{20,21}). In turn, this would also
reduce the effectiveness of potential Bd mitigation strategies. For
example, a limited ability by amphibians to develop resistance or
tolerance suggests that reintroduction of amphibians into sites to
reestablish populations extirpated by Bd will often result not in
population recovery, but instead in the rapid reinfection and mortality
of the introduced animals and/or their progeny\textsuperscript{22--25}.
If true, the future of many amphibian species threatened by Bd appears
bleak.

The Sierra Nevada yellow-legged frog (\emph{Rana sierrae}), is
emblematic of the global decline of amphibians caused by
Bd\textsuperscript{8}. Once the most common amphibian in the high
elevation portion of California's Sierra Nevada mountains,
USA\textsuperscript{26}, during the past century this frog has
disappeared from more than 90\% of its historical
range\textsuperscript{27}. Due to the severity of its decline and the
increasing probability of extinction, \emph{R. sierrae} is now listed as
``endangered'' under the U.S. Endangered Species Act. The decline of
\emph{R. sierrae} was initiated by the introduction of non-native trout
into the extensive fishless region of the Sierra
Nevada\textsuperscript{28,29} starting in the late 1800s. The arrival of
Bd in the mid-1900s and its subsequent spread\textsuperscript{30} caused
additional large-scale population extirpations\textsuperscript{31,32}.
These Bd-caused declines are fundamentally different from the
fish-caused declines because fish eradication is
feasible\textsuperscript{33} and results in the rapid recovery of frog
populations\textsuperscript{34,35}. In contrast, Bd appears to persist
in habitats even in the absence of amphibian hosts\textsuperscript{36},
and therefore represents a long-term alteration of invaded ecosystems
that amphibians will need to overcome to reestablish populations.

Despite the catastrophic impact of Bd on \emph{R. sierrae}, wherein most
Bd-naive populations are extirpated following Bd
arrival\textsuperscript{31}, some populations have persisted after
epizootics\textsuperscript{37} and are now
recovering\textsuperscript{16} (Fig.~\ref{fig-recovery-model}). During
the epizootic phase, \emph{R. sierrae} are highly susceptible to Bd
infection as indicated by very high infection intensities (``load''). In
contrast, frogs in recovering populations have reduced susceptibility to
Bd infection\textsuperscript{16}, with loads on adults consistently in
the low-to-moderate range\textsuperscript{37--39}. This reduced
susceptibility is evident even under controlled laboratory
conditions\textsuperscript{16}, indicative of host resistance against Bd
infection and not simply an effect of factors external to individual
frogs (e.g., environmental conditions). In addition to frogs from
recovering populations having higher resistance to Bd infection than
those from naive populations, they could also have higher tolerance, but
no data are currently available to evaluate this possibility. Therefore,
we focus on resistance throughout this paper.

The observed resistance of \emph{R. sierrae} could be the result of
several mechanisms that are not mutually exclusive, including natural
selection for more resistant genotypes\textsuperscript{20,21}, acquired
immunity\textsuperscript{19}, and/or inherent between-population
differences that pre-date Bd exposure. The possible evolution of
resistance and subsequent population recovery is consistent with that
expected under ``evolutionary rescue'', whereby rapid evolutionary
change increases the frequency of adaptive alleles and restores positive
population growth\textsuperscript{40,41}. This intriguing possibility
also suggests an opportunity to expand recovery beyond the spatial scale
possible under natural recovery by using resistant frogs from recovering
populations in reintroductions to vacant habitats\textsuperscript{39,42}
(Fig.~\ref{fig-recovery-model}).

@@

Bd loads were fairly consistent before versus after translocation, and
loads were nearly always well below the level indicative of severe
chytridiomycosis (i.e., the disease caused by Bd) and associated frog
mortality\textsuperscript{31,39} (Supplementary Fig. 1). Although it is
possible that the observed relatively small changes in load are a
consequence of individuals with high Bd loads dying and therefore being
unavailable for sampling during the post-translocation period, the fact
that there was little difference in pre- versus post-translocation Bd
loads even in those populations that had very high frog survival (70556,
74976 - see below; Supplementary Fig. 1) suggests a true lack of
substantial change in Bd load.

The ultimate measure of reintroduction success is the establishment of a
self-sustaining population. Given that it can take years or even decades
to determine the self-sustainability of a reintroduced population (for
an example in \emph{R. sierrae}, see\textsuperscript{39}), the use of
proxies is essential for providing shorter-term insights into
reintroduction success and the factors driving it. Results from our CMR
surveys allowed us to accurately estimate frog survival, including over
the entire CMR time series for each site and during only the 1-year
period immediately following translocation. These estimates were made
using site-specific models analyzed using the mrmr
package\textsuperscript{43}
(\url{https://snarl1.github.io/mrmr/index.html}). We use these estimates
to describe general patterns of frog survival in all translocated
cohorts, and in an among-site meta-analysis of frog survival to identify
important predictors of 1-year frog survival (e.g., Bd load).

@@

Results from the frog translocation study are suggestive of population
establishment. However, a decade or more of surveys may be necessary to
confirm that populations are in fact
self-sustaining\textsuperscript{39}. To extend our inferences of
population establishment beyond those possible from the site-specific
CMR data, we developed a population viability model. Specifically, to
test whether the observed yearly adult survival probabilities in
translocated populations were sufficient for long-term viability, we
built a stage-structured matrix model that captured known frog
demography and included demographic and environmental stochasticity. We
parameterized the model using CMR data from translocated populations and
known life history values in this system (Supplementary Table 1).
Results from a sensitivity analysis are provided in Supplementary Fig.
4.

@@

Disease-induced population declines are decimating global
biodiversity\textsuperscript{5}, but broadly-applicable strategies to
recover affected species are generally lacking
(e.g.,\textsuperscript{13}). Here, we tested the possibility that
populations of resistant individuals from naturally recovering
populations can be used to reestablish extirpated populations of the
endangered \emph{R. sierrae} in the presence of a highly virulent
pathogen (Bd). Our results indicate (i) the capacity of reintroduced
populations to become established and eventually recover despite ongoing
disease, (ii) that post-translocation frog survival is influenced by
site, cohort, and individual-level characteristics, and (iii) that 50\%
of the reintroduced populations are likely to persist over a 50-year
period. In light of the generally low success rate of amphibian
reintroduction efforts\textsuperscript{44}, our success in
reestablishing \emph{R. sierrae} populations via reintroduction of
resistant individuals is striking, and even more so given that \emph{R.
sierrae} was driven to near-extinction by Bd. Collectively, these
results provide a rare example of amphibian recovery in the presence of
Bd, and have important implications for the conservation and recovery of
amphibians and other taxa worldwide that are endangered by escalating
impacts from emerging infectious diseases.

Previous field studies in \emph{R. sierrae} show that frog-Bd dynamics
and frog survival in the presence of Bd are fundamentally different
between naive and recovering populations. Following the arrival and
establishment of Bd in previously-naive populations, adult frogs develop
high Bd loads that lead to mass die-offs\textsuperscript{31}. In
contrast, in recovering populations adult frogs typically have
low-to-moderate and relatively constant Bd loads and mass die-offs are
not observed\textsuperscript{37,38} (see also Supplementary Fig. 2). The
differences in Bd load of frogs from naive and recovering populations
are also observed in controlled laboratory studies (see Figure 4
in\textsuperscript{16}), and clearly indicate that frogs from recovering
populations exhibit resistance against Bd infection. Finally,
reintroductions of \emph{R. sierrae} collected from Bd-naive populations
and reintroduced into Bd-positive sites have consistently failed due to
the development of high Bd loads and resulting high frog mortality,
suggesting the importance of frog resistance in population
reestablishment in the presence of Bd.

@@

Results from the population viability model were also encouraging. In
particular, translocated populations with \textgreater{} 50\% survival
in the first year post-translocation were predicted to have a low
probability of extinction over 50 years (probability of extinction \(<\)
0.5 when year-1 juvenile survival probability was greater than 0.10).
The viability model highlighted the important role of frog survival in
affecting long-term population viability, and allowed us to extend the
temporal scale of our study beyond the years covered by our
post-translocation surveys. These long-term forecasts are important,
given that reintroduced \emph{R. sierrae} populations may often take
decades to achieve our ultimate goal of
self-sustainability\textsuperscript{39}. Making well-supported
projections about the long-term outcome of reintroduction efforts from
shorter-term information is crucial for adaptive management of species
reintroduction programs\textsuperscript{45}, including the one we are
carrying out for \emph{R. sierrae}. Specifically, the combined results
from our reintroduction study and viability modeling suggest that
survival of frogs in the first year following translocation is an
effective proxy of longer-term survival and population viability. In
addition, given the repeatability of frog survival at a site, 1-year
frog survival also serves as an effective proxy of site quality (i.e.,
the ability of a site to support high frog survival and a viable frog
population over the long term). This proxy of site quality is important
in the \emph{R. sierrae} system because accurately predicting the
ability of a site to support a viable frog population \emph{a priori}
remains difficult, even after conducting 24 translocations over 16
years.

@@

Despite the demonstrated resistance of adult \emph{R. sierrae} against
Bd infection, individual and population-level impacts of Bd are still
evident. In an earlier study of 2 of our 12 translocated
populations\textsuperscript{39}, Bd infection and load had detectable
effects on the survival of adults and may have influenced population
establishment (sites referred to as ``Alpine'' and ``Subalpine''
in\textsuperscript{39} are identified as ``70550'' and ``70505'' in the
current study). Applying similar analyses to all 12 of our translocated
populations would likely provide a broader perspective of the ongoing
effect of Bd. In addition to these important but relatively subtle
effects of Bd on adults, the impacts on younger life stages are more
apparent. \emph{R. sierrae} immediately following metamorphosis
(``metamorphs'') are highly susceptible to Bd
infection\textsuperscript{46} and as a result experience high
mortality\textsuperscript{32}. This high susceptibility of metamorphs is
documented in numerous species of anurans, and may result from the
poorly developed immune system characteristic of this life
stage\textsuperscript{47}. In both naturally recovering and reintroduced
\emph{R. sierrae} populations, we suggest that the high mortality of
metamorphs is an important limitation on subsequent recruitment of new
adults. Therefore, although adult \emph{R. sierrae} appear relatively
resistant, Bd infection continues to have important limiting effects on
recovering populations (see also\textsuperscript{48}). The ongoing high
susceptibility of metamorphs to chytridiomycosis also highlights the
critical need to develop better methods of monitoring this vulnerable
life stage, the results of which could significantly advance the
conservation and recovery of Bd-impacted amphibians.

Our successful use of translocations to reestablish \emph{R. sierrae}
populations despite ongoing Bd infection is particularly important
because natural recovery alone appears insufficient to allow
reestablishment of populations at the landscape scale. Naturally
recovering \emph{R. sierrae} populations in Yosemite are relatively rare
and are often isolated from other such populations and from suitable but
vacant habitats. This isolation is the consequence of both natural
fragmentation due to rugged geography and the ongoing presence of
introduced, predatory trout in the majority of lakes and
streams\textsuperscript{49}. Collectively, these factors often prevent
naturally recovering \emph{R. sierrae} populations from expanding via
dispersal of frogs to nearby unoccupied habitats. Translocations allowed
us to overcome these dispersal limitations and reestablish populations
in sites that would otherwise remain unoccupied for the foreseeable
future. Of particular note are the translocations that we conducted as
part of the current study into areas from which \emph{R. sierrae} were
completely extirpated, but instead of single isolated habitats these
areas contained habitats composed of interconnected fishless lakes,
ponds, marshes, and/or streams. Successful reestablishment of one
\emph{R. sierrae} population in such areas has produced a source of
frogs from which adults and juveniles are now dispersing into adjacent
habitats. This process of translocation, establishment, and subsequent
dispersal has significant potential to eventually allow the
reestablishment of robust metapopulations that may be more resilient to
current and future natural and anthropogenic impacts than more
fragmented populations (e.g.,\textsuperscript{50}).
